D3064R1: How to Avoid OOTA Without Really Trying

### 2024-05-05 ootavacuous.tex
Update paper number and date
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Appendix D.4 ("Inventing Atomic Loads"): Apply Alan Stern feedback
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{document}
\title{D3064R0: How to Avoid OOTA Without Really Trying}

@@

\author{
Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
Maged Michael\\{\small \email{maged.michael@gmail.com}}
}
\date{April 5, 2024 (Post-Pre-Tokyo)}
\maketitle{}

@@

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.

@@

>>>>

\begin{document}
\title{D3064R1: How to Avoid OOTA Without Really Trying}

@@

\author{
Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
Maged Michael\\{\small \email{maged.michael@gmail.com}}
}
\date{May 4, 2024 (Pre-St. Louis)}
\maketitle{}

@@

However, before making such a choice, we should look at other situations.

\subsubsection{Customary Non-Atomic-Load ``Invention''}
\label{app:Customary Non-Atomic-Load Invention}

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.
Hence the use of quotation marks in the section heading:
Here the load is not actually invented, but instead merged with an
earlier load with no synchronization between the two loads.

@@

In addition, compiler writers should carefully consider the issues
raised by the following section:

\subsubsection{Other Cautions for Atomic-Load Invention}
\label{app:Other Cautions for Atomic-Load Invention}

Consider the code below:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
A developer might be forgiven for assuming that the value of \co{r1}
is guaranteed to be a perfect square, at least for sufficiently small
values of \co{x}.

But if the compiler is permitted to invent atomic loads, there is no such
guarantee, as can be seen from this transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was ten, then the value of \co{r1} will be twenty, which is not
a perfect square.

Similar issues can arise even when the compiler merely repurposes a
value from an atomic load, as can be seen from this example:
\begin{quote}
\begin{verbatim}
do_something(x);
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
Again, a developer might be forgiven for assuming that the value of
\co{r1} is guaranteed to be a perfect square, at least for sufficiently
small values of \co{x}.
But is the compiler allowed to transform this example into the following?
\begin{quote}
\begin{verbatim}
repurposed = x;
do_something(repurposed);
int r0 = x;
int r1 = r0 * r0 + 2 * repurposed + 1;
\end{verbatim}
\end{quote}
Again, this can result in the value of \co{r1} being something other than
a perfect square.
%
But it would be just fine if the compiler fully merged the two loads from
\co{x}, for example, as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
do_something(r0);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the developer expected line~19 to pass two fields from the
same structure to \co{do_something_with()}, which is an entirely
reasonable expectation.
This expectation will not always be met if the compiler is permitted
to invent atomic loads as shown by the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free,
the second load on line~15 could be removed without violating user
expectations.

It is imperative that users be able to rely on arithmetic and pointer accesses.
Therefore, compilers must refrain from inventing atomic loads and from repurposing
their values, but compilers are permitted to merge adjacent atomic loads from the
same variable.
These restrictions correspond to our notion of quasi volatile defined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
